---
title: "Flow Analysis Tutorial"
author: "Jolie Phan"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 4
  html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

```{r}
library(here)
library(CytoML)
library(flowCore)
library(flowWorkspace)
library(COMPASS)
```

# Load data

The .xml and .fcs files for this dataset are stored in the Seshadri Lab shared 
drive (LSR Fortessa/2021 Summer HAARVIVAC/20210902 HAARVIVAC Batch 4/).

Download the .xml file and folder containing the .fcs files and drag into the 
"data" folder of this project directory.

```{r}
# Location of XML file
xml_path <- here::here("data/20211014_HAARVIVAC_B4V3_JP.xml")

# Location of .fcs files
fcs_subfolder <- here::here("data/20210902_HAARVIVAC_FCS_B4/")
```

Create a flowjo_workspace object with the function open_flowjo_xml().

```{r}
ws <- open_flowjo_xml(xml_path)
```

# Create a GatingSet

## Set-up

A GatingSet holds a set of GatingHierarchy objects, representing a set of samples 
and the gating scheme associated with each.

Look at the workspace metadata to choose which keywords to extract into the 
GatingSet. The flowjo_to_gatingset() function parses a flowJo Workspace to 
generate a GatingSet object.

```{r}
names(fj_ws_get_keywords(ws, 117)) 
keywords2import <- c("EXPERIMENT NAME",
                       "$DATE",
                       "SAMPLE ID",
                       "PATIENT ID",
                       "Stim",
                       "WELL ID",
                       "PLATE NAME") 
sampleGroup <- "Samples"

gs <- flowjo_to_gatingset(ws,                                    
                          name = sampleGroup, 
                          keywords = keywords2import,
                          path = fcs_subfolder, 
                          extend_val = -10000)
```

## QC

Make sure that the gating trees are consistent for all samples.

```{r}
pop_lists <- lapply(gs, gh_get_pop_paths)
unique(pop_lists)
```

Remove channels from flow data that are not used by gates.

```{r}
gs <- gs_remove_redundant_channels(gs) # drop SSC-H, V655-A, V570-A
```

Add names to all channels or fix their names.

```{r}
dput(unname(pData(parameters(gh_pop_get_data(gs[[1]])))[,2]))
markernames <- c("Time", "FSC-A", "FSC-H", "SSC-A", "CD8b", "TNFa", "CD107a",
                 "CD154", "CD3", "IL2", "CD4", "IL17a", "IL4_5_13", "CD14_19",
                 "CCR7", "CD38", "LD", "IFNg", "CD45RA", "HLADR")
names(markernames) <- pData(parameters(gh_pop_get_data(gs[[1]])))[,1]
markernames(gs) <- markernames
pData(parameters(gh_pop_get_data(gs[[1]])))[,c(1,2)]
```

Plot gating tree.

```{r, fig.width = 8, fig.height = 6}
plot(gs, fontsize=15, bool=T)
```

## Save GatingSet

This can take a while.

```{r}
if(!dir.exists(here::here("out/GatingSet"))) {
  cat(sprintf("Creating folder %s\n", here::here("out/GatingSet")))
  dir.create(here::here("out/GatingSet"), recursive = T)
}
save_gs(gs, here::here("out/GatingSet"))
```

# Create a COMPASSContainer

A COMPASSContainer is the data structure used to hold data from an ICS experiment.
The input for this code is a GatingSet or a GatingSetList. Counts, metadata, and 
single cell data are extracted and fed into the COMPASSContainer constructor.

```{r}
# A regular expression to match a single node in the gating tree
parent_node <- "CD4+"

# A character identifying the subject id column in the GatingSet metadata
id <- "SAMPLE ID"

# markermap contains the output of markernames(gs)
markernames(gs) 
markermap <- list("IL2", "IL4_5_13", "IFNg", "TNFa", "IL17a", "CD154", "CD107a")
names(markermap) <- paste0(parent_node, "/", c("IL2+", "IL4_5_13+", "IFNg+",
                                               "TNFa+", "IL17a+", "CD154+", 
                                               "CD107a+"))
```

Construct the COMPASSContainer.

```{r}
CC <- COMPASSContainerFromGatingSet(gs,
                                    node = parent_node,
                                    individual_id = id,
                                    mp = markermap,
                                    countFilterThreshold = 5000)
```

Look at some basic info about our COMPASSContainer.

```{r}
CC
```

# Run COMPASS

Fit the COMPASS model using the COMPASSContainer. To fit the COMPASS model, we 
need to specify how to identify the samples that are our treatment condition and 
our control condition. Here, we will run COMPASS on the samples stimmed by spike
1 with DMSO as our negative control. For now, let's just do 100 iterations for 
speed.

```{r}
fit <- COMPASS(CC,
               treatment = Stim == "S1",
               control = Stim == "DMSO",
               iterations = 100)
```

Save the COMPASS run output.

```{r}
if(!dir.exists(here::here("out/COMPASSResult"))) {
  cat(sprintf("Creating folder %s\n", here::here("out/COMPASSResult")))
  dir.create(here::here("out/COMPASSResult"), recursive = T)
}
saveRDS(fit, file.path("out/COMPASSResult", "COMPASSResult.rds"))
```

Save the Functionality and Polyfunctionality Scores.

```{r}
FS <- FunctionalityScore(fit)
FS_df <- data.frame(tmp = names(FS), FS = FS)
colnames(FS_df) <- c("SAMPLE ID", "FS")

PFS <- PolyfunctionalityScore(fit)
PFS_df <- data.frame(tmp = names(PFS), PFS = PFS)
colnames(PFS_df) <- c("SAMPLE ID", "PFS")

FS_PFS_df <- merge(FS_df, PFS_df, by = "SAMPLE ID")
write.table(FS_PFS_df,
            file = file.path("out/COMPASSResult", "FS_PFS.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
```

# Visualize

Plot a heatmap of the mean probability of response.

```{r, fig.width = 8, fig.height = 10}
plot(fit, show_rownames = TRUE)
```
