---
title: "Flow Analysis COMPASS Tutorial"
author: "Jolie Phan, Emma Bishop"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
editor_options:
  chunk_output_type: console
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Directory structure
To directly use the code in this tutorial, you should create an RStudio project and set up your project directory as follows.

The `data/` folder should contain all `.xml` and `.fcs` files and the `out/` folder will contain the GatingSet and all COMPASS outputs.

```{r dir_structure, echo=FALSE, message=FALSE}
# You don't need to run this! See README or html for figure.
library(DiagrammeR)
grViz(diagram = "digraph flowchart {
  node [fontname = arial, shape = retangle, fontsize=8, height=0.1]
  dir1 [label = '@@1']
  dir2 [label = '@@2']
  dir3 [label = '@@3']
  dir4 [label = '@@4']
  dir5 [label = '@@5']
  dir6 [label = '@@6']
  dir7 [label = '@@7']
  dir1 -> dir2,dir3;
  dir2 -> dir4,dir5;
  dir3 -> dir6,dir7;
  }
  
  [1]: 'Rproject/'
  [2]: 'data/'
  [3]: 'out/'
  [4]: '.xml'
  [5]: '.fcs files'
  [6]: 'GatingSet/'
  [7]: 'COMPASSResult/'
  ")
```

You can achieve this directory structure by running the following commands:

```{r, message=FALSE}
if(!dir.exists(here::here("data"))) {
  cat(sprintf("Creating folder %s\n", here::here("data")))
  dir.create(here::here("data"), recursive = T)
}
if(!dir.exists(here::here("out"))) {
  cat(sprintf("Creating folder %s\n", here::here("out")))
  dir.create(here::here("out"), recursive = T)
}
if(!dir.exists(here::here("out/GatingSet"))) {
  cat(sprintf("Creating folder %s\n", here::here("out/GatingSet")))
  dir.create(here::here("out/GatingSet"), recursive = T)
}
if(!dir.exists(here::here("out/COMPASSResult"))) {
  cat(sprintf("Creating folder %s\n", here::here("out/COMPASSResult")))
  dir.create(here::here("out/COMPASSResult"), recursive = T)
}
```

# Load libraries

```{r, message=FALSE}
library(here)
library(CytoML)
library(flowCore)
library(flowWorkspace)
library(COMPASS)
library(tidyverse)
```

# Load data
Download the .xml file and folder containing the associated .fcs files (see the README for updated directions). Then, drag them into the "data" folder of the project directory.

This data has been gated in FlowJo v9 and will be parsed using flowWorkspace.

**FYI:** When using FlowJo v9, the FlowJo workspace must be exported as an .xml file to create a flowjo_workspace object with the function open_flowjo_xml(). However, when using FlowJo v10, the FlowJo workspace can be loaded directly as a .wsp file using the same function open_flow_xml().

```{r}
# Location of XML file
xml_path <- here::here("data/20200605_COVID_ICS-B3-trunc.xml")

# Location of .fcs files
fcs_subfolder <- here::here("data/20200605_COVID_ICS-B3-FCS-trunc/")
```

Create a flowjo_workspace object with the function open_flowjo_xml().

```{r}
ws <- open_flowjo_xml(xml_path)
```

# Create a GatingSet

## Set-up

A GatingSet holds a set of GatingHierarchy objects, representing a set of samples 
and the gating scheme associated with each.

Look at the workspace metadata to choose which keywords to extract into the 
GatingSet. The flowjo_to_gatingset() function parses a flowJo Workspace to 
generate a GatingSet object.

```{r}
# Look at all of the keywords
names(fj_ws_get_keywords(ws, 117)) 
# Choose which keywords to keep
keywords2import <- c("EXPERIMENT NAME",
                       "$DATE",
                       "SAMPLE ID",
                       "PATIENT ID",
                       "STIM",
                       "WELL ID",
                       "PLATE NAME") 
sampleGroup <- "Samples"

gs <- flowjo_to_gatingset(ws,                                    
                          name = sampleGroup, 
                          keywords = keywords2import,
                          path = fcs_subfolder, 
                          extend_val = -10000)
```

## QC

Make sure that the gating trees are consistent for all samples.

```{r}
pop_lists <- lapply(gs, gh_get_pop_paths)
unique(pop_lists)
```

Remove channels from flow data that are not used by gates.

```{r}
gs <- gs_remove_redundant_channels(gs) # drop SSC-H
```

Add names to all channels or change their names.

```{r}
dput(unname(pData(parameters(gh_pop_get_data(gs[[1]])))[,2]))
markernames <- c("Time", "FSC-A", "FSC-H", "SSC-A", "CD8b", "TNFa", "CD107a",
                 "CD154", "CD3", "IL2", "CD4", "IL17a", "IL4_5_13", "CD14_19",
                 "CCR7", "CD38", "LD", "IFNg", "CD45RA", "HLADR")
names(markernames) <- pData(parameters(gh_pop_get_data(gs[[1]])))[,1]
markernames(gs) <- markernames
pData(parameters(gh_pop_get_data(gs[[1]])))[,c(1,2)]
```

In this tutorial, we will only run COMPASS on CD4+ T cells, but what if we wanted to also run COMPASS on CD8+ T cells? Currently, most of the gates are missing from the "8+" node. Let's grab the missing gates from the "NOT4+" node and add them under the "8+" node.

```{r}
gates_to_copy <- c("/Time/LD-3+/1419-3+/S/Lymph/NOT4+/107a",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/154", 
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/IL2",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/IL17",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/IL4513",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/TNF",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/CCR7+",
                   "/Time/LD-3+/1419-3+/S/Lymph/NOT4+/CD45RA+")

for(path in gates_to_copy) {
  gs_pop_add(gs, lapply(gs, gh_pop_get_gate, y=path),
             parent = "/Time/LD-3+/1419-3+/S/Lymph/8+")
}
recompute(gs, "/Time/LD-3+/1419-3+/S/Lymph/8+")
```

Next, let's get rid of the "NOT4+" node and all of its descendants.

```{r}
gs_pop_remove(gs, "/Time/LD-3+/1419-3+/S/Lymph/NOT4+")
```

Plot the gating tree.

```{r, fig.width = 6, fig.height = 4}
plot(gs, fontsize=15, bool=T)
```

## Save GatingSet

**Note:** this can take a while. 

```{r}
save_gs(gs, here::here("out/GatingSet"))
```


# Get tabular data from the GatingSet

The following steps were inspired from advice in the COMPASS [issue #81](https://github.com/RGLab/COMPASS/issues/81#issuecomment-2484533481) thread.


```{r}
# Load GatingSet if needed
gs <- load_gs(here::here("out/GatingSet"))
```


## Get matrices of expression data

Extract metadata and take a look

```{r}
metadata <- pData(gs) %>%
  mutate(sampleName = rownames(.))

head(metadata)
```

Separate the Spike 1 stimmed and DMSO samples

```{r}
sp1_samples <- metadata %>%
  filter(STIM == "Spike 1") %>%
  pull(sampleName)
spike1 <- gs[sp1_samples]

dmso_samples <- metadata %>%
  filter(STIM == "DMSO") %>%
  pull(sampleName)
dmso <- gs[dmso_samples]
```

Extract data for the "CD4+" population only. These are 'cytoset' objects, which are a collection of 'cytoframe's.

```{r}
spike1_cd4 <- gs_pop_get_data(spike1, "4+")  # Using "+4" node name
dmso_cd4 <- gs_pop_get_data(dmso, "4+")
```

We can see the sample names in the object like so:

```{r}
sampleNames(dmso_cd4)
```

Let's take a look at one of these 'cytoframes', which are similar to a matrix of expression values.

```{r}
samplename <- "114716.fcs_445737"
cytoframe1 <- dmso_cd4[[samplename]]
head(cytoframe1)
```

Note that we have gates we don't need for our boolean combinations, like time and forward/side scatter (Time, FSC-A, FSC-H, SSC-A). We will get rid of these below.

Get a list of expression matrices for stim and unstim from the cytoset/cytoframe objects.

```{r}
markernames <- c("Time", "FSC-A", "FSC-H", "SSC-A", "CD8b", "TNFa", "CD107a",
                 "CD154", "CD3", "IL2", "CD4", "IL17a", "IL4_5_13", "CD14_19",
                 "CCR7", "CD38", "LD", "IFNg", "CD45RA", "HLADR")

markermap <- list("IL2", "IL4_5_13", "IFNg", "TNFa", "IL17a", "CD154", "CD107a")

get_mtx_list <- function(cytoset_obj) {
  outlist = list()
  for (i in sampleNames(cytoset_obj)) {
    sub_cf <- get_cytoframe_from_cs(cytoset_obj, i)  # Extract cytoframe from cytoset
    out <- exprs(sub_cf)  # Get expression values as a matrix
    colnames(out) <- markernames  # Rename columns (e.g. <V450-A> becomes IFNg)
    out[out < 0] <- 0  # Replace negative numbers with 0 for CellCounts() later
    out <- out[, colnames(out) %in% markermap]  # Only keep markers we want boolean combos of
    outlist[[i]] <- out
  }
  return(outlist)
}

spikelist <- get_mtx_list(spike1_cd4)
dmsolist <- get_mtx_list(dmso_cd4)
```

Now we have named lists of matrices of expression data, using sample names as indices

```{r}
names(dmsolist)
```

```{r}
head(dmsolist[[1]])
```

Save

```{r}
saveRDS(spikelist, here::here("out/spikelist.rds"))
saveRDS(dmsolist, here::here("out/dmsolist.rds"))
```

## Create boolean combinations

```{r}
spikelist <- readRDS(here::here("out/spikelist.rds"))
dmsolist <- readRDS(here::here("out/dmsolist.rds"))
```

Can take a little while

```{r}
# Seven markers
n_s <- CellCounts(spikelist, Combinations(7)) # Stimmed data
n_u <- CellCounts(dmsolist, Combinations(7)) # Unstimmed data
```

Save count matrices and metadata

```{r}
saveRDS(n_s, here::here("out/n_s.rds"))
saveRDS(n_u, here::here("out/n_u.rds"))
saveRDS(metadata, here::here("out/metadata.rds"))
```


# Run SimpleCOMPASS

```{r}
n_s <- readRDS(here::here("out/n_s.rds"))
n_u <- readRDS(here::here("out/n_u.rds"))
metadata <- readRDS(here::here("out/metadata.rds"))
```

```{r}
# Use subject ID as rownames
rownames(n_s) <- c("23", "25", "133C", "142C", "150C")
rownames(n_u) <- c("23", "25", "133C", "142C", "150C")
```

Fits a polyfunctionality response model to our data.

```{r, message=FALSE}
fit = COMPASS::SimpleCOMPASS(n_s = n_s,
                             n_u = n_u,
                             meta = metadata,
                             individual_id = "SAMPLE ID")
```

Save the COMPASS run output.

```{r}
saveRDS(fit, file.path(here::here("out/COMPASSResult"), "COMPASSResult.rds"))
# fit <- readRDS(file.path(here::here("out/COMPASSResult"), "COMPASSResult.rds"))
```

Save the Functionality and Polyfunctionality Scores.

```{r}
FS <- FunctionalityScore(fit)
FS_df <- data.frame(tmp = names(FS), FS = FS)
colnames(FS_df) <- c("sampleName", "FS")

PFS <- PolyfunctionalityScore(fit)
PFS_df <- data.frame(tmp = names(PFS), PFS = PFS)
colnames(PFS_df) <- c("sampleName", "PFS")

FS_PFS_df <- merge(FS_df, PFS_df, by = "sampleName")
write.table(FS_PFS_df,
            file = file.path(here::here("out/COMPASSResult"), "FS_PFS.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)
FS_PFS_df
```

# Visualize

Plot a heatmap of the mean probability of response.

```{r, fig.width = 6, fig.height = 4, message=FALSE}
plot(fit, show_rownames = TRUE)
```

