---
title: "Seshadri Lab tSNE Tutorial"
author: "Nathan Kieswetter"
date: "12/21/2021"
output: html_document
---
# Introduction

[t-distributed stochastic neighbor embedding (t-SNE)](https://www.jmlr.org/papers/volume9/vandermaaten08a/vandermaaten08a.pdf) is a non-linear dimensionality reduction statistical analysis useful for assessing high-dimensional data. Specifically, it models each high-dimensional object by a two- or three-dimensional point in such a way that similar objects are modeled by nearby points and dissimilar objects are modeled by distant points with high probability. Similarly, [Uniform Manifold Approximation and Projection (UMAP)](https://umap-learn.readthedocs.io/en/latest/) is another more novel dimensionality reduction technique. Whilst competitive with t-SNE, UMAP is claimed to better preserve the global structure of the data and is less computationally intensive. For a more detailed comparison of two techniques, we recommend this [article](https://towardsdatascience.com/tsne-vs-umap-global-structure-4d8045acba17).

In flow cytometry, t-SNE and umap analyses are useful for segregating and identifying unique cellular populations based on their surface protein expression (SS) as well as the cytokines or chemokines being secreted by specific subsets as assessed via intracellular staining (ICS).

For this tutorial, we will be analyzing a batch of data from the HAARVI cohort. This guide aims to enable you to read in, QC, analyze, and visualize surface staining and intracellular staining cell data from a typical flow cytometry experiment with the assumtpion that the data has been correctly compensated and formated into an R-friendly syntax.

**NOTE**: If you have completed the Seshadrilab [COMPASS tutorial](https://github.com/seshadrilab/Flow_Analysis_Tutorial), the same (saved) gating set can be used for this tutorial, allowing you to skip the initial reading-in and QC of the raw FACS files.

# Installation

The following R packages are required for this tutorial:
-here
-flowCore
-flowWorkspace
-openCyto
-CytoML
-tidyverse

Check if the required packages are installed and install them if not:

```{r eval=, warning=FALSE, include=FALSE}
if (!require(Rtsne)) install.packages("Rtsne")
if (!require(uwot)) install.packages("uwot")
if (!require(devtools)) install.packages("devtools")
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(scales)) install.packages("scales")
if (!require(patchwork)) install.packages("patchwork")
if (!require(hues)) install.packages("hues")
if (!require(RColorBrewer)) install.packages("RColorBrewer")
if (!require(tidyselect)) install.packages("tidyselect")
if (!require(ggrastr)) install.packages("ggrastr")
if (!require(here)) install.packages("here")
if (!require(BiocManager)) install.packages("BiocManager")
if (!require(CytoML)) BiocManager::install("CytoML")
if (!require(flowCore)) BiocManager::install("flowCore")
if (!require(flowWorkspace)) BiocManager::install("flowWorkspace")
```

# Load in the Appropriate Libaries

```{r results ='hide', message = FALSE, results = FALSE, warning = FALSE}
library(here)
library(flowCore)
library(flowWorkspace)
library(openCyto)
library(CytoML) 
library(tidyverse)
library(Rtsne)
library(devtools)
library(scales)
library(patchwork)
library(hues)
library(RColorBrewer)
library(tidyselect)
library(ggrastr)
theme_set(theme_bw(18))
```

# Load in the Data

```{r}
date <- 20211220
### Location of XML file
xml_path <- here::here("data/20211014_HAARVIVAC_B4V3_JP.xml")

### Location of .fcs files
fcs_subfolder <- here::here("data/20210902_HAARVIVAC_FCS_B4/")
```

Create a flowjo_workspace object with the function open_flowjo_xml().

```{r}
ws <- CytoML::open_flowjo_xml(xml_path)
```

The .xml and .fcs files for this dataset are stored in the Seshadri Lab shared drive (LSR Fortessa/2021 Summer HAARVIVAC/20210902 HAARVIVAC Batch 4/). Download the .xml file and folder containing the .fcs files and drag into the "data" folder of this project directory.

# Create a GatingSet
### Set-up

A GatingSet holds a set of GatingHierarchy objects, representing a set of samples and the gating scheme associated with each. Look at the workspace metadata to choose which keywords to extract into the GatingSet. The *flowjo_to_gatingset()* function parses a flowJo Workspace to generate a GatingSet object.

```{r}
names(fj_ws_get_keywords(ws, 117)) 
```

Select the keywords of interest; i.e., those which contain useful information.

```{r}
keywords2import <- c("EXPERIMENT NAME",
                     "$DATE",
                     "SAMPLE ID",
                     "PATIENT ID",
                     "Stim",
                     "WELL ID",
                     "PLATE NAME") 
sampleGroup <- "Samples"
```

Create the gatingset.

```{r}
gs <- flowjo_to_gatingset(ws,                                    
                          name = sampleGroup, 
                          keywords = keywords2import,
                          path = fcs_subfolder, 
                          extend_val = -10000)
```

### Quality Control (QC)

Ensure that the gating trees are consistent for all samples.

```{r}
pop_lists <- lapply(gs, gh_get_pop_paths)
unique(pop_lists)
```

Remove channels from flow data that are not used by gates.

```{r}
gs <- gs_remove_redundant_channels(gs) 
```

Add or fix all channel names.

```{r}

dput(unname(pData(parameters(gh_pop_get_data(gs[[1]])))[,2]))

markernames <- c("Time", "FSC-A", "FSC-H", "SSC-A", "CD8b", "TNFa", "CD107a",
                 "CD154", "CD3", "IL2", "CD4", "IL17a", "IL4_5_13", "CD14_19",
                 "CCR7", "CD38", "LD", "IFNg", "CD45RA", "HLADR")

names(markernames) <- pData(parameters(gh_pop_get_data(gs[[1]])))[,1]

markernames(gs) <- markernames

pData(parameters(gh_pop_get_data(gs[[1]])))[,c(1,2)]
```

Plot the gating tree. Make sure there are no errant nodes etc.

```{r}
plot(gs, fontsize=15, bool=T)
```

### Save the gatingset

Note, this can take a while.

```{r include = TRUE, eval = FALSE}
if(!dir.exists(here::here("out/GatingSet"))) {
  cat(sprintf("Creating folder %s\n", here::here("out/GatingSet")))
  dir.create(here::here("out/GatingSet"), recursive = T)
}
save_gs(gs, here::here("out/GatingSet"))

```



# Preparation of Data for t-SNE 

With the data read-in, QC'ed and converted into a gating set, it is time to extract the data required for downstream dimred analysis.

### Define the populations from which you like to extract their counts 

These populations can be pulled up using: dput(gh_get_pop_paths(gs))

```{r}
pops2extract <- c("/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/CCR7+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/CD45RA+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/CD107a+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/CD154+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/IFNg+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/IL2+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/IL4_5_13+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/IL17a+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/TNFa+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/CCR7+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/CD45RA+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/CD107a+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/CD154+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/IFNg+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/IL2+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/IL4_5_13+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/IL17a+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/TNFa+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD38+", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/DN", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/DP", 
                  "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/HLADR+")
```

### Extract the Counts

The following chunk of code extracts the count data for the populations defined above. Additionally, it renames the populations of interest for easier visualization of the raw data within a given data frame or table. Use *dplyr::select()* to define useful columns for your analyses. These should include columns containing patient identifiers, treatments, etc. 

```{r}
pop_counts <- gs_pop_get_count_with_meta(gs, subpopulations = pops2extract) %>%
  dplyr::select(`PATIENT ID`, Stim, sampleName, Population, Count) %>%
  mutate(Population = gsub(".*Lymphocytes\\/", "", Population)) %>%
  mutate(Population = recode(Population,
                             "Singlet/Live" = "T_cells", 
                             "Singlet/Live/CD4+" = "CD4", 
                             "Singlet/Live/CD4+/CCR7+" = "CD4_CCR7", 
                             "Singlet/Live/CD4+/CD45RA+" = "CD4_CD45RA", 
                             "Singlet/Live/CD4+/CD107a+"= "CD4_CD107", 
                             "Singlet/Live/CD4+/CD154+" = "CD4_CD154", 
                             "Singlet/Live/CD4+/IFNg+" = "CD4_IFNy", 
                             "Singlet/Live/CD4+/IL2+" = "CD4_IL2", 
                             "Singlet/Live/CD4+/IL4_5_13+" = "CD4_IL4_5_13", 
                             "Singlet/Live/CD4+/IL17a+" = "CD4_IL17", 
                             "Singlet/Live/CD4+/TNFa+" = "CD4_TNFa", 
                             "Singlet/Live/CD8+" = "CD8", 
                             "Singlet/Live/CD8+/CCR7+" = "CD8_CCR7", 
                             "Singlet/Live/CD8+/CD45RA+" = "CD8_CD45RA", 
                             "Singlet/Live/CD8+/CD107a+" = "CD8_CD107a", 
                             "Singlet/Live/CD8+/CD154+" = "CD8_CD154", 
                             "Singlet/Live/CD8+/IFNg+" = "CD8_IFNg", 
                             "Singlet/Live/CD8+/IL2+" = "CD8_IL2", 
                             "Singlet/Live/CD8+/IL4_5_13+" = "CD8_IL4_5_13", 
                             "Singlet/Live/CD8+/IL17a+" = "CD8_IL17", 
                             "Singlet/Live/CD8+/TNFa+"= "CD8_TNF", 
                             "Singlet/Live/CD38+"= "CD38_pos", 
                             "Singlet/Live/DN" = "DN", 
                             "Singlet/Live/DP" = "DP", 
                             "Singlet/Live/HLADR+" = "HLADR_pos")) %>% 
  pivot_wider(id_cols = c(`PATIENT ID`, Stim, sampleName),
              names_from = Population,
              values_from = Count)
```

These data can (and should) be saved for quick analysis later on.

### Sample the Data

In some instances, it may be prudent to sample your data so that each condition (i.e. "Stim"ulation) has an equal number of cells or observations. This can reduce computational burden as well as prevent potential bias; however, it may not be necessary for all analyses. Here, the CD4 cell population was selected and each "Stim" sampled 5000 times (i.e. 5000 cells per "Stim" per patient).

Below is a custom function which randomly samples sub-groups and distribute these data for downstream analyses. 
```{r results ='hide', message = FALSE, results = FALSE}
distributeEvents <- function(totalEvents, subGroupSizes) {
  if(totalEvents == 0) {
    rep(0, length(subGroupSizes))
  } else if(sum(subGroupSizes) < totalEvents) {
    stop("Not enough events to achieve requested sample size")
  } else if(any(subGroupSizes == 0)) {
    output <- rep(0, length(subGroupSizes))
    nonZeroIndices <- which(subGroupSizes > 0)
    output[nonZeroIndices] <- distributeEvents(totalEvents, subGroupSizes[nonZeroIndices])
  } else {
    numSubGroups <- length(subGroupSizes)
    total_div_numSubGroups <- totalEvents %/% numSubGroups
    remainder <- totalEvents %% numSubGroups
    minGroupSize <- min(subGroupSizes)
    
    if(total_div_numSubGroups == 0) {
      stopifnot(remainder != 0)
      stopifnot(all(subGroupSizes > 0))
      stopifnot(remainder < numSubGroups)
      output <- rep(0, numSubGroups)
      output[sample.int(numSubGroups, remainder)] <- 1
      output
    } else {
      howMuchToAssign <- min(total_div_numSubGroups, minGroupSize)
      output <- rep(howMuchToAssign, numSubGroups)
  
      remaining_subGroupSizes <- subGroupSizes - howMuchToAssign
      nonZeroIndices <- which(remaining_subGroupSizes > 0)
      if(length(nonZeroIndices)) {
        toAdd <- rep(0, numSubGroups)
        toAdd[nonZeroIndices] <- distributeEvents(totalEvents - sum(output), remaining_subGroupSizes[nonZeroIndices])
        output + toAdd
      } else {
        output
      }
    }
  }
}
```

Sample the CD4 data by "Stim"

```{r}
cd4_tsne_sample_sizes <- pop_counts %>% 
  dplyr::select(`PATIENT ID`, Stim, sampleName, CD4) %>% 
  group_by(Stim) %>% 
  nest() %>% 
  ungroup() %>% 
  mutate(nsamp = map2(5000, data, function(totalEvents, df) {
    distributeEvents(totalEvents, df$CD4)
  })) %>% 
  unnest(cols = c(data, nsamp))

cd4_tsne_sample_sizes %>% 
  group_by(Stim) %>% 
  summarise(nsamp_sum = sum(nsamp)) %>% 
  dplyr::pull(nsamp_sum) %>% 
  unique()
```

### Create a data matrix to input into t-SNE/UMAP

First, define the gates that you like to include in the matrix and include them in a new object. These data can be viewed and selected using the gh_get_pop_paths() function.

```{r}

dput(gh_get_pop_paths(gs))

gates_to_extract_membership <- c("/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/CCR7+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/CD45RA+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/CD107a+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/CD154+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/IFNg+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/IL2+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/IL4_5_13+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/IL17a+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/TNFa+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/CCR7+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/CD45RA+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/CD107a+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/CD154+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/IFNg+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/IL2+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/IL4_5_13+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/IL17a+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD8+/TNFa+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD38+", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/DN", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/DP", 
                                 "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/HLADR+")
```

Next, assess whether the defined populations (*gates_to_extract_membership()*) exist in the gating set. This is a useful QC step at this stage of the analysis to ensure that the data being associated is correct.

The result should be returned as: **TRUE**

```{r}
all(gates_to_extract_membership %in% intersect(gh_get_pop_paths(gs), gh_get_pop_paths(gs)))
```

Prepare the matrix for the sampled CD4 cells. Is is important to define the parent gate for the chunk of code below (i.e. "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+").

Below is a custom function to sample n events from a GatingHierarchy containing a single sample.

```{r results ='hide', message = FALSE, results = FALSE}
sampleGatingHierarchy <- function(gh, parentGate, n, otherGates = NULL) {
  stopifnot(length(gh) == 1)
  allMarkerNames <- pData(parameters(gh_pop_get_data(gh)))[,c(1,2)]
  if (any(is.na(allMarkerNames[,2])) | length(unique(allMarkerNames[,2])) < length(allMarkerNames[,2]))
    stop ("all marker names (even FSC-A and Time) must be assigned and be unique")
  availableEvents <- gh_pop_get_count(gh, parentGate)
  nSampled <- sample.int(availableEvents, size = n)
  parentGateIndices <- gh_pop_get_indices(gh, parentGate)
  gates2Extract <- unique(c(parentGate, if(is.null(otherGates)) {gh_get_pop_paths(gh)} else {otherGates}))
  perCellGateMembership <- data.frame(lapply(gates2Extract, function(currentGate) {
    as.integer(gh_pop_get_indices_mat(gh, currentGate)[parentGateIndices,][nSampled]) }))
  colnames(perCellGateMembership) <- gates2Extract
  perCellMFIData <- exprs(gh_pop_get_data(gh, parentGate))[nSampled,]
  colnames(perCellMFIData) <- allMarkerNames[match(colnames(perCellMFIData), allMarkerNames[,1]), 2]
  stopifnot(nrow(perCellGateMembership) == nrow(perCellMFIData))
  gateAndMFIData <- cbind(perCellGateMembership, perCellMFIData)
  cbind(pData(gh), gateAndMFIData, row.names = NULL)
}

```

Extract the CD4 specific count and membership data and input it into the prepared matrix.

```{r}
cd4_data4dimred <- map2_dfr(cd4_tsne_sample_sizes$sampleName, cd4_tsne_sample_sizes$nsamp,
                            function(currentSampleName, currentSampleSize) {
                              # print(sprintf("Sampling data from %s", currentSampleName))
                              sampleGatingHierarchy(gs[[currentSampleName]],
                                                    "/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+",
                                                    currentSampleSize,
                                                    gates_to_extract_membership)
                            })
dim(cd4_data4dimred)
cd4_data4dimred[1:4,]
```

## Scale the data and apend it to the CD4 matrix

Scaling centers and/or scales the columns of a numeric matrix by the normalization of a dataset using the mean value and standard deviation and is a standard step in many analyses. We highly recommend scaling all data prior to further analysis.

```{r}
pData(parameters(gh_pop_get_data(gs[[1]])))[,c(1,2)] 

dimred_input_cols <- c("FSC-A", "FSC-H", "SSC-A", "CD8b", "TNFa", "CD107a", 
                       "CD154", "CD3", "IL2", "CD4", "IL17a", "IL4_5_13", 
                       "CD14_19", "CCR7", "CD38", "LD", "IFNg", "CD45RA", 
                       "HLADR")

#CD4 Scaled Data for rTSNE
cd4_data_scaled <- cd4_data4dimred %>% 
  dplyr::select(all_of(dimred_input_cols)) %>% 
  as.matrix() %>% 
  scale() %>% 
  as.data.frame() %>% 
  rename_at(vars(all_of(dimred_input_cols)),function(x) paste0(x,".scaled"))

cd4_data4dimred_scaled <- bind_cols(cd4_data4dimred, cd4_data_scaled)
```

# Run t-SNE

The following steps may take a significant amount of time depending on the number of variables, computational power, and the number of cells sampled. Further, the number of iterations can also be altered which can also have a significant effect on analysis time.


### Set the seed

A seed is a necessary numerical marker used by t-SNE analyses to ensure reproducibility. If not defined, R will randomly assign a seed and the t-SNE/UMAP output may vary from run to run. Below, we have used the date on which the following tutorial was created; however, this can be any value you want.

```{r}
set.seed(date)
```
Define the common markers that are present in all the populations you which to assess.
```{r}
common_tsne_cols <- c("CD3", "CD4", "CD8b", 
                      "CD38", "CD4", "HLADR", 
                      "CD45RA")
```

Define the marker (ICS or SS) that you would like to include into the t-SNE analysis, and run:
```{r}
print(Sys.time())
CD4_tsne_out <- cd4_data4dimred_scaled %>%
  dplyr::select(all_of(paste0(c(common_tsne_cols, "TNFa", "CD107a", "CD154", "IL2", 
                                                  "IL17a", "IL4_5_13","CCR7", "CD38", 
                                                  "IFNg", "CD45RA", "HLADR"), ".scaled"))) %>% 
  Rtsne::Rtsne(check_duplicates = FALSE)
```

Append the t-SNE data to your original matrix and save this file.

```{r}
cd4_tsne_and_data_scaled <- cbind(as.data.frame(CD4_tsne_out$Y) %>% 
                                    dplyr::rename(x.tsne = V1, y.tsne = V2),
                                  cd4_data4dimred)
```

# Basic plotting

The following code will plot the t-SNE data in two dimensions. Facet_grid() is a useful function for stratifying your data by treatment (i.e. "Stim").

```{r}
cd4_tsne_and_data_scaled %>%
  ggplot(aes(x = x.tsne, 
             y = y.tsne,
             color = Stim)
  ) +
  geom_point() +
  theme(legend.position="FALSE") +
  facet_grid(vars(Stim))
```

# Advanced Plotting

The ability to overlay ICS or SS data onto the t-SNE plot we generated above is a powerful tool that can allow for the generation of biological hypotheses and conclusions.

The following chunk of code defines a set theme from which the proceeding plots are generated. This can be easily altered to your individual needs.

```{r}
myPalette <- colorRampPalette(rev(brewer.pal(11, "Spectral")))

pointSize <- 0.002

run_tmp <- cd4_tsne_and_data_scaled

gg_themes <- ggplot(run_tmp, aes(x = x.tsne, y = y.tsne)) +
  scale_x_continuous(limits = range(run_tmp$x.tsne)*1.01) +                   
  scale_y_continuous(limits = range(run_tmp$y.tsne)*1.01) +                     
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size=22, face="bold"),
        axis.text = element_blank(),
        axis.line = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_blank(),
        plot.margin = margin(1, 0, 0, 0, unit = "pt"),
        strip.text = element_text(size=12))
```

Below, is an example whereby TNFa expression is overlaid onto the CD4 t-SNE data and then stratified (as.factor(x)) by stimulation condition. This can be repeated for any surface marker or stained cytokine by altering the code accordingly.

Specific populations can be easily called using  dput(gh_get_pop_paths(gs))
```{r}
TNFa_plot <- gg_themes +
  geom_point(aes(color=as.factor(`/Time/CD3+/CD14-CD19-/Lymphocytes/Singlet/Live/CD4+/TNFa+`)),
             shape=20, alpha=0.9, size=pointSize, show.legend = FALSE) +
  scale_color_manual(values = c("0" = "#E2E2E2", "1" = "#023FA5")) +
  theme(legend.position = NULL) +
  guides(colour = guide_legend(override.aes = list(size=9))) +
  ggtitle("TNFa") +
  labs(color = "TNFa")+
  facet_grid(vars(Stim))

TNFa_plot
```



